JD notes on effects

### 2022-07-11 manuscript/body.tex
Little beak :( after intro

To make an outcome plot, we use a \emph{focal} predictor on the x-axis. The resulting plot will depend on choices we make about other (non-focal) predictor(s). In ordinary (Gaussian) linear regression, the non-focal choices may have a simple additive effect on the outcome estimate. However, when the focal predictor has interactions or a non-linear link function, non-focal choices can also affect the slope of the response. Additional challenges arise when dealing with “mixed” models, which incorporate random effects.

@@

Generating quantities, i.e., central estimate together with the corresponding confidence intervals has some challenges. In particular:
\begin{enumerate}
\item choosing the \emph{reference point} for non-focal predictors in multivariate models
\item uncertainty estimation -- appropriate choice of \emph{anchor} for computing confidence intervals in effects plots
\item \jd{When is this a challenge? Doesn't it just work like magic with the lm machinery? and how to incorporate the uncertainty due to non-focal predictors}
\item biases induced by non-linear transformations of the response variable in generalized models (especially generalized mixed models).
\item choice of representative values of the focal predictor \jd{I still don't see this as much of a challenge.}
\end{enumerate}

Representative values for a focal predictor are generally chosen using quantiles, or equally spaced values, for a continuous predictor; or an exhaustive set of levels for a categorical predictor.
The central estimates are then calculated by holding the non-focal predictors at a reference point while varying the focal predictor, with the goal that the estimates represent how the model responds to the changes in the focal predictor \citep{fox2009effect, hanmer2013behind}. These values have been called: \emph{predictor effects} \citep{fox2009effect}, \emph{marginal predictions} \citep{leeper2017package} or \emph{estimated marginal means} \citep{lenth2018package} \jd{Is this also the same as least-squares means?}. In this article, we refer to these quantities as the \emph{central estimates} of the outcome.

\jd{It looks like you're trying to introduce column averaging and non-linear averaging in the same \P. How can we simplify?}
For linear models (including models with complex interactions), the averaging is done on the linear scale, i.e., linear averaging. As a result, the \emph{bias} in the expected (mean) predictions can easily be corrected using \emph{mean-based reference point}, i.e., averaging of the non-focal model variables -- currently not implemented in commonly used \proglang{R} software packages. The mean of the central estimates is the prediction at the \emph{model center}. In this case, the estimates are usually consistent with the observed values. However, in models with non-linear link functions, this is not always the case. In particular, when dealing with GL(M)Ms with non-linear link functions, the averaging is done on the non-linear scale, i.e., non-linear averaging, making generating correct predictions much harder. This is the bias in the expected mean prediction induced by the non-linear transformation of the response variable. One way to address this is to make predictions on the linear predictor scale and back-transform to the original scale. However, the back-transformation may either result in biased predictions or requires some approximation, for example, second-order Taylor approximation implemented in \pkg{emmeans} \citep{lenth2018package}. An alternative to the mean-based reference point is the \emph{whole-sample-based} approach, discussed in detail later, which involves computing the prediction over the population of non-focal predictors and then averaging across the values of the focal predictor \citep{hanmer2013behind}.

This article aims to discuss and implement various approaches for computing predictions and effects, together with the associated plots. We further explore and demonstrate, using simulated data, approaches for correcting bias in central estimates for GL(M)Ms involving non-linear link functions. The proposed method and \proglang{R} software package will complement the existing ones by providing: 1) a straightforward way to generate effects, and 2) an alternative and a more robust way to correct for prediction bias in GL(M)Ms.

@@

\jd{I guess this list should really be a desc, not an itemize.}

\begin{itemize}
\item \textbf{Input variables}: Refers to the observed (or scientific) variables underlying an inference or exploration. For example, the regression models described by \EREF{simple_inter_higher_no_interaction} and \EREF{simple_inter_higher} both have $3$ input variables -- $x_1, x_2, x_3$.
\item \textbf{Focal predictors}: We call the input variable on the x-axis of an outcome plot the focal predictor. 
Any other predictor variables are “non-focal” predictors. 
For example, consider the model described by \EREF{simple_inter_higher_no_interaction}. 
If we are focusing on the effect of $x_1$ on the response $y$, then $x_1$ is the focal predictor while $x_2$ and $x_3$ are non-focal predictors. \jd{Is this trivial example necessary? Is any example necessary here?}

\item \textbf{Model matrix}: Refers to the design matrix whose rows include all combination of input variables. Consider an example for three hypothetical households -- the first household head is a Christian with an income of $\$ 50$, the second household head is Muslim with an income of $\$ 100$ while the third household head is a Jew with an income of $\$ 77$. Suppose we want to model the household size (\code{hhsize}) as function of these household characteristics, i.e., $$\mathrm{hhsize} = \beta_0 + \beta_1\times\mathrm{income} + \beta_{2[r]}\times\mathrm{religion} + \mathrm{error}.$$ The model matrix corresponding to this model is given by
$$\begin{bmatrix}{}
 Intercept & income & religionJew & religionMuslim \\
 1 & 50 & 0 & 0 \\
  1 & 100 & 0 & 1 \\
  1 & 77 & 1 & 0 \\
\end{bmatrix}.$$ The first column represents the constant term in our model, $\beta_0$. For continuous input variables, the representation in the model matrix is the same as the corresponding input variables (for example second column representing income). For categorical variables, however, by default, the model matrix  creates additional dummy variables using the reference cell parameterization. This means that, if an input variable has $L$ factor levels, then there will be $L-1$ dummy columns representing all but the first level created in the model matrix. In our example, the column \code{religion} is missing and instead we have \code{religionJew} and \code{religionMuslim}; the missing category \code{religionChristian} is treated as the reference category.

\item \textbf{Model variables}: Refer to the variables which are combined to make the linear predictor (corresponding to the columns in the model matrix). Each input variable may correspond to one or more model variable. Input variables with more than two categories (for instance religion in our previous example), or input variables with non-linear response functions (e.g., spline or polynomial) will correspond to more than one model variable -- we call such variables “multi-parameter variables”.

\item \textbf{Model center:} A point corresponding to a column-wise mean of the model matrix (the mean of one or more model variables). Also referred to as center point. The center point for a set of model variables corresponding to an input variable may not represent a possible value of the input variable -- the case in \FREF{justify_plots}B. In simple linear models with no interactions, the mean of the input variable can be used as a proxy for the center point \jd{Is it a “proxy” or is it the thing??}. If there are more than one model variables associated with the focal input variable, however, the model center does not correspond to any given value of the focal predictor \jd{This seems like bad repetition}, and we do not necessarily get the ``nice'' crossing we get in simpler cases, see \FREF{pred_cubic_plots}.

@@

\item \textbf{Marginal effects:} \jd{This isn't really one of our definitions; maybe some of this belongs in the text part.} In simple linear models with no interaction terms, the coefficient estimates are simple and directly interpretable as the expected change in outcome for a unit change in focal input variable. This is the \emph{unconditional marginal effect} and it is constant across all the observations and levels of all other input variables. Consider models described by \EREF{simple_inter_higher_no_interaction} and \EREF{simple_inter_higher}. The marginal effect of $x_2$ in model~\ref{eq:simple_inter_higher_no_interaction} is $\frac{\partial y}{\partial x_2} = \beta_2$. On the other hand, the marginal effect of $x_2$ in model~\ref{eq:simple_inter_higher} is given by $\frac{\partial y}{\partial x_2} = \beta_2 + \beta_{23}x_3$. In other words, if there are no interactions, the marginal effect of $x_2$ on $y$ is constant, while, if there are interactions in the model, the marginal effect of a change in $x_2$ on $y$ depends on the value of the other \emph{conditioning} predictor, $x_3$. In essence, computation of marginal effects requires the use of partial derivatives and can be considered as the slope of a multi-dimensional surface with respect to the surface of interest \cite{leeper2017interpreting}.

\item \jd{Should we consider defining effects-style and prediction-style CIs? Can we think of better words?}

@@

One may be interested in the uncertainties associated with the focal predictor only, excluding other uncertainties due to other non-focal predictors -- effect plots. However, as previously mentioned, the two packages do not provide a straightforward way do achieve this. To illustrate this, we consider \EREF{simple_inter_higher_no_interaction} and use $x_2$ as the focal predictor, and generate the outcome plot shown in \FREF{justify_ci_plots}. 
%
\begin{figure}
\begin{center}
\includegraphics{justify_anchors_all.ggp.pdf}
\end{center}
\caption{{\bf Prediction and effect plots.} The description of horizontal, vertical and trend lines remain the same as above. A: The wider dashed curves corresponds to the conventional prediction curves from \pkg{emmeans} and \pkg{effects} packages, while the narrower curves crossing at the mean of the focal predictor, also referred to as the center point, represent the effect from \pkg{varpred}. For simple OLS models, the effect-styled curves cross at the center point (which is the default anchor and also referred to as center-anchored). The prediction-styled curves incorporates not only the uncertainties due to intercept term but also other non-focal predictors. On the other hand, effect-styled curves only take into account the main effect uncertainty of the focal predictor. B: center- and zero- anchored effects. The center-anchored effects (as described above) while the zero-anchored means that the anchor is at zero value of the focal predictor. The choice of the anchor do not affect the central estimates, hence they are identical in both choices of the anchor.}
\label{fig:justify_ci_plots}
\end{figure}
%
For \pkg{emmeans} and \pkg{effects}, the confidence intervals are much wider because they include uncertainties associated with the intercept and non-focal predictors, but narrower and crosses at the mean of the focal predictor in \pkg{varpred}. In other words, with \pkg{varpred}, we are able to generate effects indicating zero uncertainty at the value of the focal predictor we are more certain about, i.e., the anchor. For simple OLS models, the point where the curves cross is the model center or simply center point; it corresponds to anchor we choose, in this case, the center point (which is generally an appropriate and stable choice). Predictions are the standard curves; they tell us what we expect to see if we know the value of the focal predictor. The lines inside (effects) are a little different; they are a good way to visualize the effect of the focal predictor. Unlike predictions, where the uncertainty captures the uncertainty in the whole model, the effects plot focuses on uncertainty in the effect of the focal predictor only and depend on the anchor.

>>>>

To make an outcome plot, we use a \emph{focal} predictor on the x-axis and central estimate (predicted values) on the y-axis. The resulting plot will depend on choices we make about other (non-focal) predictor(s). In ordinary (Gaussian) linear regression, the non-focal choices may have a simple additive effect on the central estimate. However, when the focal predictor has interactions or a non-linear link function, non-focal choices can also affect the slope of the estimate. Additional challenges arise when dealing with “mixed” models, which incorporate random effects.

@@

Generating quantities, i.e., central estimate together with the corresponding confidence intervals has some challenges. In particular:
\begin{enumerate}
\item choosing the \emph{reference point} for non-focal predictors in multivariate models
\item uncertainty estimation -- appropriate choice of \emph{anchor} for computing confidence intervals in effects plots
\item \jd{When is this a challenge? Doesn't it just work like magic with the lm machinery? and how to incorporate the uncertainty due to non-focal predictors} \BC{Clarify with JD. Another challenge or a continuation  of the above?} 
\item biases induced by non-linear transformations of the response variable in generalized models (especially generalized mixed models).
\item choice of representative values of the focal predictor \jd{I still don't see this as much of a challenge.} \dya{We needed to mention somewhere that we also have to choose focal values. I think \P~ below is suffecient.}
\end{enumerate}

Representative values for a focal predictor are generally chosen using quantiles, or equally spaced values, for a continuous predictor; or an exhaustive set of levels for a categorical predictor.
The central estimates are then calculated by holding the non-focal predictors at a reference point (a value chosen for a non-focal predictor) while varying the focal predictor, with the goal that the estimates represent how the model responds to the changes in the focal predictor \citep{fox2009effect, hanmer2013behind}. These values have been called: \emph{predictor effects} \citep{fox2009effect}, \emph{marginal predictions} \citep{leeper2017package} or \emph{estimated marginal means} \citep{lenth2018package} \jd{Is this also the same as least-squares means?} \BC{Yes. But was later renamed to emmeans}. In this article, we refer to these quantities as the \emph{central estimates} of the outcome.

In models with non-focal predictors such as multivariate models, reference point can be chosen as the average of the non-focal \emph{model variables} -- we call this approach \emph{mean-based} reference point and is currently not implemented in commonly used \proglang{R} software packages. The mean of the central estimates is the prediction at the \emph{model center}. We introduce an alternative choice for the reference point below.

For linear models, the averaging is done on the linear scale, i.e., linear averaging. As a result, the \emph{bias} in model center estimates can easily be corrected using mean-based approach and the estimates are usually consistent with the observed values. However, in models with non-linear link functions, this is not always the true. In particular, when dealing with generalized models with non-linear link functions, the averaging is done on the non-linear scale, i.e., non-linear averaging, making generating correct predictions much harder. This is the bias in the expected mean prediction induced by the non-linear transformation of the response variable. One way to address this is to make predictions on the linear predictor scale and back-transform to the original scale. However, the back-transformation may either result in biased predictions or requires some approximation, for example, second-order Taylor approximation implemented in \pkg{emmeans} \citep{lenth2018package}. An alternative to the mean-based reference point is the \emph{whole-sample-based} approach, discussed in detail later, which involves computing the prediction over the population of non-focal predictors and then averaging across the values of the focal predictor \citep{hanmer2013behind}.

This article aims to discuss and implement various approaches for computing predictions and effects, together with the associated plots. We further explore and demonstrate, using simulated data, approaches for correcting bias in central estimates for generalized (mixed) models involving non-linear link functions. The proposed method and \proglang{R} software package will complement the existing ones by providing: 1) a straightforward way to generate effects, and 2) an alternative and a more robust way to correct for prediction bias in generalized (mixed) models.

@@

\jd{I guess this list should really be a desc, not an itemize.}\BC{Clarify}

\begin{itemize}
\item \textbf{Input variables:} Refers to the observed (or scientific) variables underlying an inference or exploration. For example, the regression models described by \EREF{simple_inter_higher_no_interaction} and \EREF{simple_inter_higher} both have $3$ input variables -- $x_1, x_2, x_3$.
\item \textbf{Focal predictors}: We call the input variable on the x-axis of an outcome plot the focal predictor.  Any other predictor variables are ``non-focal'' predictors.

\item \textbf{Model matrix:} Refers to the design matrix whose rows include all combination of input variables. Consider an example for three hypothetical households -- the first household head is a Christian with an income of $\$ 50$, the second household head is Muslim with an income of $\$ 100$ while the third household head is a Jew with an income of $\$ 77$. Suppose we want to model the household size (\code{hhsize}) as function of these household characteristics, i.e., $$\mathrm{hhsize} = \beta_0 + \beta_1\times\mathrm{income} + \beta_{2[r]}\times\mathrm{religion} + \mathrm{error}.$$ The model matrix corresponding to this model is given by
$$\begin{bmatrix}{}
 Intercept & income & religionJew & religionMuslim \\
 1 & 50 & 0 & 0 \\
  1 & 100 & 0 & 1 \\
  1 & 77 & 1 & 0 \\
\end{bmatrix}.$$ The first column represents the constant term in our model, $\beta_0$. For continuous input variables, the representation in the model matrix is the same as the corresponding input variables (for example second column representing income). For categorical variables, however, by default, the model matrix  creates additional dummy variables using the reference cell parameterization. This means that, if an input variable has $L$ factor levels, then there will be $L-1$ dummy columns representing all but the first level created in the model matrix. In our example, the column \code{religion} is missing and instead we have \code{religionJew} and \code{religionMuslim}; the missing category \code{religionChristian} is treated as the reference category.

\item \textbf{Model variables:} Refer to the variables which are combined to make the linear predictor (corresponding to the columns in the model matrix). Each input variable may correspond to one or more model variable. Input variables with more than two categories (for instance religion in our previous example), or input variables with non-linear response functions (e.g., spline or polynomial) will correspond to more than one model variable -- we call such variables “multi-parameter variables”.

\item \textbf{Model center:} A point corresponding to a column-wise mean of the model matrix (the mean of one or more model variables). Also referred to as center point. The center point for a set of model variables corresponding to an input variable may not represent a possible value of the input variable -- the case in \FREF{justify_plots}B. In simple linear models with no interactions, the mean of the input variable can be used is a center point. If there are more than one model variables associated with the focal input variable, however, the model center does not correspond to a single value of the focal predictor, see \FREF{pred_cubic_plots}.

@@

\item \textbf{Prediction-style and effect-style plots:} A prediction-style plot is about predicting observations for a given value of the focal predictor. For this, we want to use the classic curved confidence intervals. An effect-style plot attempts to visualize the effect of a focal predictor and are characterized with narrower confidence intervals. Unlike a prediction-style plot, where the confidence intervals capture the uncertainty associated all predictors in the model, the effect-style plot focuses on uncertainty associated with the focal predictor only and depend on the anchor.

@@

One may be interested in the uncertainties associated with the focal predictor only, excluding other uncertainties due to other non-focal predictors -- effect plots. However, as previously mentioned, the two packages do not provide a straightforward way do achieve this. To illustrate this, we consider \EREF{simple_inter_higher_no_interaction} and use $x_2$ as the focal predictor, and generate the outcome plot shown in \FREF{justify_ci_plots}. 
%
\begin{figure}
\begin{center}
\includegraphics{justify_anchors_all.ggp.pdf}
\end{center}
\caption{{\bf Prediction and effect plots.} The description of horizontal, vertical and trend lines remain the same as above. A: The wider dashed curves corresponds to the conventional prediction curves from \pkg{emmeans} and \pkg{effects} packages, while the narrower curves crossing at the mean of the focal predictor, also referred to as the center point, represent the effect from \pkg{varpred}. For simple OLS models, the effect-styled curves cross at the center point (which is the default anchor and also referred to as center-anchored). The prediction-styled curves incorporates not only the uncertainties due to intercept term but also other non-focal predictors. On the other hand, effect-styled curves only take into account the main effect uncertainty of the focal predictor. B: center- and zero- anchored effects. The center-anchored effects (as described above) while the zero-anchored means that the anchor is at zero value of the focal predictor. The choice of the anchor do not affect the central estimates, hence they are identical in both choices of the anchor.}
\label{fig:justify_ci_plots}
\end{figure}
%
For \pkg{emmeans} and \pkg{effects}, the confidence intervals are much wider because they include uncertainties associated with the intercept and non-focal predictors, but narrower and crosses at the mean of the focal predictor in \pkg{varpred}. In other words, with \pkg{varpred}, we are able to generate effects indicating zero uncertainty at the value of the focal predictor we are more certain about, i.e., the anchor. For simple OLS models, the point where the curves cross is the model center or simply center point; it corresponds to anchor we choose, in this case, the center point (which is generally an appropriate and stable choice).

\BC{To clean%
\textbf{Marginal effects:} \jd{This isn't really one of our definitions; maybe some of this belongs in the text part.} In simple linear models with no interaction terms, the coefficient estimates are simple and directly interpretable as the expected change in outcome for a unit change in focal input variable. This is the \emph{unconditional marginal effect} and it is constant across all the observations and levels of all other input variables. Consider models described by \EREF{simple_inter_higher_no_interaction} and \EREF{simple_inter_higher}. The marginal effect of $x_2$ in model~\ref{eq:simple_inter_higher_no_interaction} is $\frac{\partial y}{\partial x_2} = \beta_2$. On the other hand, the marginal effect of $x_2$ in model~\ref{eq:simple_inter_higher} is given by $\frac{\partial y}{\partial x_2} = \beta_2 + \beta_{23}x_3$. In other words, if there are no interactions, the marginal effect of $x_2$ on $y$ is constant, while, if there are interactions in the model, the marginal effect of a change in $x_2$ on $y$ depends on the value of the other \emph{conditioning} predictor, $x_3$. In essence, computation of marginal effects requires the use of partial derivatives and can be considered as the slope of a multi-dimensional surface with respect to the surface of interest \cite{leeper2017interpreting}.%
}
